Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-03 TACO-paper-v2/coffee.tex
Add explanation of linear elasticity
Some other minor adjustments

\begin{abstract}
We study and systematically evaluate a class of composable code 
transformations that improve arithmetic intensity in local assembly 
operations, which represent a significant fraction of the execution
time in finite element methods. Their performance optimization is indeed a 
challenging issue. Even though affine loop nests are generally present, 
the short trip counts and the complexity of mathematical expressions, which 
vary among different problems, make it hard to determine a single or unique 
sequence of successful transformations. Our investigation has resulted in the 
implementation of a compiler for local assembly kernels, fully
integrated with a real framework for developing finite element methods. 
The compiler manipulates abstract syntax trees generated from a domain-specific
language for partial differential equations by introducing domain-aware optimizations,
tailored to instruction-level parallelism and register locality. Eventually, C code including 
vector SIMD intrinsics is produced. Experiments using a range of finite-element 
problems of increasing complexity show that significant performance improvement 
is achieved. The generality of the approach and the applicability of the proposed 
code transformations to other domains is also discussed.

@@

Table~\ref{table:code-transformations} summarizes the code
transformations described so far. Given that many of these
transformations depend on some parameters (e.g. tile size), we need a
mechanism to prune such a large optimization space. This aspect is
treated in Section~\ref{sec:pyop2-compiler}.

@@

The compiler executes an ordered sequence of optimization steps to the 
ASTs received from FFC. Application of \emph{licm} must precede padding 
and data alignment, due to the introduction of temporary arrays. These
transformations are always performed because they are likely to improve
the run-time performance, as demonstrated by our results in Section~\ref{sec:analysis-code-transf}.
Based on a cost model (described later), loop interchange, \emph{split}, 
and \emph{op-vect} may be introduced. Their implementation is centered 
on analysis and manipulation of the kernel's AST. When \emph{op-vect}
is selected, the compiler outputs AVX or AVX-512 intrinsics code. Any 
possible corner cases are handled: for example, if \emph{op-vect} is to 
be applied, but the size of the iteration space is not a multiple of the vector 
length, then a remainder loop, amenable to auto-vectorization, is inserted.

@@

We first observe that manual full (or extensive) unrolling is unlikely to be effective. The \texttt{ijk} loop nest would need to be small enough such that the unrolled instructions do not exceed the instruction cache, which is rarely the case: it is true that in a local assembly kernel the minimum size of the \texttt{ijk} loop nest is 3$\times$3$\times$3 (triangular mesh and polynomial order 1), but this increases rapidly with the polynomial order of the method and the discretization employed (e.g. tetrahedral meshes imply larger loop nests than triangular ones), so sizes greater than 10$\times$10$\times$10, for which extensive unrolling would already be harmful, are in practice very common. In addition, manual unrolling is dangerous because it may compromise compiler auto-vectorization by either removing loops (most compilers search for vectorizable loops) or losing spatial locality within a vector register.

@@

\subsection{Full-application Study}
\label{sec:full-app}
In this section, we investigate the performance gain for an entire finite element computation
developed in Firedrake, a static linear elasticity problem. The source code is available at~\cite{elasticity-code}. The application simulates ... .

@@

\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. Throughout the paper we assumed to assign the execution of a kernel to a single thread, which is likely to be the best strategy for a standard multi-core CPU. On the other hand, we note that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability. 
\item The various sub-expressions in $\beta$ only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.  
\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation concerns resource-aware execution. It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; nevertheless, the basic idea is far more general, which is re-scheduling (re-associating) operations to optimize for some generic parameters. It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by less stringent memory requirements (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}). In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also the evaluation of $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression. This may be particularly indicated for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.

@@

\item \textbf{Code generation and applicability of the transformations.} All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time. This means that ``good'' code can be generated. For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable. Further, all of these factors contribute to the applicability and the effectiveness of all of our code transformations. For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}). On the other hand, the padding and data alignment transformation could be performed directly at run-time; note that, theoretically, it could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.  
\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become effective for a variety of codes. Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge. However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results). Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved. The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.

@@

>>>>

\begin{abstract}
We study and systematically evaluate a class of composable code 
transformations that improve arithmetic intensity in local assembly 
operations, which represent a significant fraction of the execution
time in finite element methods. Their performance optimization is indeed a 
challenging issue. Even though affine loop nests are generally present, 
the short trip counts and the complexity of mathematical expressions, which 
vary among different problems, make it hard to determine a single or unique 
sequence of successful transformations. Our investigation has resulted in the 
implementation of a compiler for local assembly kernels, fully
integrated with a framework for developing finite element methods. 
The compiler manipulates abstract syntax trees generated from a domain-specific
language by introducing domain-aware optimizations and, eventually,
produces C code including vector SIMD intrinsics. Experiments using a range of finite-element 
problems of increasing complexity show that significant performance improvement 
is achieved. The generality of the approach and the applicability of the proposed 
code transformations to other domains is also discussed.

@@

Table~\ref{table:code-transformations} summarizes the code
transformations described so far. Given that many of these
transformations depend on some parameters, we need a
mechanism to prune such a large optimization space, which can be as large as order of thousands variants for medium-size loops and moderately complex mathematical expressions (all parametric transformations can indeed be combined). This aspect is treated in Section~\ref{sec:pyop2-compiler}.

@@

The compiler applies an ordered sequence of optimization steps to the 
ASTs received from FFC. Application of \emph{licm} must precede padding 
and data alignment, due to the introduction of temporary arrays. These
transformations are always performed because they are likely to improve
the run-time performance, as demonstrated by our results in Section~\ref{sec:analysis-code-transf}.
Based on a cost model (described later), loop interchange, \emph{split}, 
and \emph{op-vect} may be introduced. Their implementation is centered 
on analysis and manipulation of the kernel's AST. When \emph{op-vect}
is selected, the compiler outputs AVX or AVX-512 intrinsics code. Any 
possible corner cases are handled: for example, if \emph{op-vect} is to 
be applied, but the size of the iteration space is not a multiple of the vector 
length, then a remainder loop, amenable to auto-vectorization, is inserted.

@@

We first observe that manual full (or extensive) unrolling is unlikely to be effective for two reasons. Firstly, the \texttt{ijk} loop nest would need to be small enough such that the unrolled instructions do not exceed the instruction cache, which is rarely the case: it is true that in a local assembly kernel the minimum size of the \texttt{ijk} loop nest is 3$\times$3$\times$3 (triangular mesh and polynomial order 1), but this increases rapidly with the polynomial order of the method and the discretization employed (e.g. tetrahedral meshes imply larger loop nests than triangular ones), so sizes greater than 10$\times$10$\times$10, for which extensive unrolling would already be harmful, are in practice very common. Secondly, manual unrolling is dangerous because it may compromise compiler auto-vectorization by either removing loops (most compilers search for vectorizable loops) or losing spatial locality within a vector register.

@@

\subsection{Full-application Study}
\label{sec:full-app}
In this section, we investigate the performance gain for an entire finite element computation
developed in Firedrake, a static linear elasticity problem. The source code is available at~\cite{elasticity-code}. The equation is used to simulate deformation of an object caused by pre-established loading conditions.

@@

\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. Throughout the paper we assumed a kernel be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs. On the other hand, we note that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability. 
\item The various sub-expressions in $\beta$ only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises. Pre-computation of invariant terms also increases memory requirements due to the need for temporary arrays, so it is possible that for certain architectures the transformation could actually cause slow downs (e.g. whenever the available per-core memory is small). 
\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation concerns resource-aware execution. It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; nevertheless, the basic idea is far more general, which is re-scheduling (re-associating) operations to optimize for some generic parameters. It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by less stringent memory requirements (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}). In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression. This may be particularly indicated for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.

@@

\item \textbf{Code generation and applicability of the transformations.} All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time. This means that ``good'' code can be generated. For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable. Further, all of these factors contribute to the applicability and the effectiveness of some of our code transformations. For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}). A special case are padding and data alignment since they could be performed even at run-time if some values were not known at code generation time; theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.  
\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become increasingly effective in a variety of codes. Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers, to the best of our knowledge. The outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization. In addition, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi experiments). Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that this could be more easily achieved in specific domains. The intuition is to map the memory access pattern on vector registers, and then exploiting in-register shuffling to minimize the traffic between memory and processor. By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way. 

@@

%We conclude noting that despite experiments were performed on Intel platforms only, performance gains are expected even on completely different CPU architectures, since the techniques proposed in this paper involves re-scheduling
